Expert-In-The-Loop Causal Discovery: \\ Iterative Model Refinement Using Expert Knowledge

### 2025-02-10 paper/main.tex
UPdates

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters like $X$, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ is the number of variables in the set. We denote the
covariance between variables $ X $ and $ Y $ as $ \mathrm{cov}(X, Y) $, a 
covariance matrix as $ \Sigma $, and the entry corresponding to the covariance
between $ X $ and $ Y $ as $ \Sigma_{XY} $. We consider causal DAGs that contain 
only observed variables. A DAG $ G = (V, E) $ is an acyclic directed graph
whose nodes $ V $ corresponding to random variables and whose edges $ E $ represent
direct causal relationships \cite{Pearl2009}. For an edge $e=X \to Y$, 
$\textrm{reverse}(e)=Y \to X$. The set of parents of $ X $ in $ G $
is denoted as $ \textrm{Pa}_G(X) $, and its ancestors and descendants as $
\textrm{An}_G(X) $ and $ \textrm{De}_G(X) $, respectively; we use the convention
that $X \notin \textrm{An}_G(X)$ and $X \notin \textrm{De}_G(X)$. We denote a path, $
\pi(X, Y) = \{ X, V_0, V_1, \cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G
$ where consecutive pairs of variables in $ \pi $ are connected by an
edge. We say $ X $ and $ Y $ are d-connected in $ G $ given a conditioning set
$ \bm{Z} \subseteq V - \{X, Y\} $, if for a path $ \pi(X, Y) $ (i) for every
collider structure ($ V_1 \rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $
\bm{Z} \cap \{ V_2, \textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every
non-collider structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1
\leftarrow V_2 \rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are
\emph{d-separated} given $ \bm{Z} $ if no d-connecting path exist between them.

@@

\begin{figure}
	\centering
	\begin{tabular}{llrr}
		$X$ & $Y$ & $ \bm{Z} $ & p-value \\
		\hline
		Incm & MrtS &  Age, Occp, Rltn & $ 0.29 $     \\
		Incm & Sex  &  Occp, Rltn      & $ 0.21 $     \\
		Incm & Wrkc &  Occp 	       & $ 1.5e^{-3} $ \\
		Edct & MrtS & Incm, Occp, Rltn & $ 0.02 $       \\
		\hline
	\end{tabular}
	\caption{Results of testing some of implied CIs of the DAG in
		 Figure~\ref{fig:intro}(b) using the same CI test that was used for
		 learning it.}
	\label{fig:ci_table}
\end{figure}

@@

whereas a \emph{weak ancestral oracle} is defined as:
$$
\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{rand}(\}X \gets Y, X \to Y\}) & \textrm{otherwise} \\
\end{cases}
$$
where $\textrm{rand}(S)$ denotes a uniformly chosen element of $S$.
That is, our ancestral oracles only provide information on the direction of cause-effect
relationships, and do not distinguish between direct and indirect effects. 
The weak ancestral oracle is even unable to detect the absence of cause-effect
relationships between two variables. Using such oracles for DAG construction  
inherently risks that superfluous edges are created.

@@

In this section, we compare our \textsc{ExpertInLoop} algorithm with automated
causal discovery algorithms. Our goal with the empirical analysis is to
understand the behaviour of our method when the experts are not perfect. To
simulate an imperfect domain expert, we use a version of the strong ancestral oracle
that has an accuracy of $ \alpha $:
\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		\mathrm{Expert}(\alpha) &= \begin{cases} 
			\mathcal{O}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
			\mathcal{O}(X, Y) &= \begin{cases}
				X \rightarrow Y, & \textrm{if } X \rightarrow Y \in G_{\textrm{true}} \\
				Y \rightarrow X, & \textrm{if } Y \rightarrow X \in G_{\textrm{true}} \\
					\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}
where $G_{\textrm{true}}$ is the true data generating DAG.

It is important to note that the effective accuracy of $
\mathrm{Expert} $ is higher than $ \alpha $ as even when giving a random
answer, there is a $ 1 $ in $ 3 $ chance that the answer is correct. Therefore,
an $ \mathrm{Expert} $ with accuracy $ \alpha $, has an effective accuracy, $ \alpha_{\mathrm{eff}} = \alpha + (1 - \alpha) / 3 $.

@@

To compare the learned DAG to the original DAG, we use two metrics: Structural
Hamming Distance (SHD) and Structural Intervention Distance (SID)
\citep{Peters2015}. Unlike \textrm{ExpertInLoop}, the automated algorithms can
only recover the MEC, therefore, we considered all possible orientations of the
MEC to compute a range of SHD and SID values representing the best and worst
case scenarios. The results of this analysis are shown in
Figure~\ref{fig:shd_sid}. For SHD, the performance of \textsc{ExpertInLoop} is
comparable to the automated algorithms for $ \alpha = 0.3
(\alpha_{\textrm{eff}} = 0.53 $, implying that if an expert is able to get the
edge orientation correctly in more than $ 1 $ out of $ 2 $ cases, they can
outperform the automated algorithms. Similarly for SID, an expert can outperform 
the automated algorithms if they are able to get the orientation correct in $ 2 $ 
out of $ 3 $ cases. Another thing to note is that for denser DAGs, the 
\textsc{ExpertInLoop} is able to perform better for lower $ \alpha $ values.

>>>>

\section{Background}
\label{sec:background}
We denote random variables using uppercase letters like $X$, and a set of
random variables by $ \bm{X} = \{X_1, \cdots, X_k\} $, where $ \rvert \bm{X}
\rvert = k $ is the number of variables in the set. We denote the
covariance between variables $ X $ and $ Y $ as $ \mathrm{cov}(X, Y) $, a 
covariance matrix as $ \Sigma $, and the entry corresponding to the covariance
between $ X $ and $ Y $ as $ \Sigma_{XY} $. We consider causal DAGs that contain 
only observed variables. A DAG $ G = (V, E) $ is an acyclic directed graph
whose nodes $ V $ corresponding to random variables and whose edges $ E $ represent
direct causal relationships \cite{Pearl2009}. For an edge $e=X \to Y$, 
$\textrm{reverse}(e)=Y \to X$. The set of parents of $ X $ in $ G $
is denoted as $ \textrm{Pa}_G(X) $, and its ancestors and descendants as $
\textrm{An}_G(X) $ and $ \textrm{De}_G(X) $, respectively; we use the convention
that $X \notin \textrm{An}_G(X)$ and $X \notin \textrm{De}_G(X)$. We denote a path, $
\pi(X, Y) = \{ X, V_0, V_1, \cdots, V_k, Y \} $ between $ X $ and $ Y $ in $ G
$ where consecutive pairs of variables in $ \pi $ are connected by an
edge. We say $ X $ and $ Y $ are d-connected in $ G $ given a conditioning set
$ \bm{Z} \subseteq V - \{X, Y\} $, if for a path $ \pi(X, Y) $ (i) for every
collider structure ($ V_1 \rightarrow V_2 \leftarrow V_3 $) on $ \pi(X, Y) $, $
\bm{Z} \cap \{ V_2, \textrm{De}_G(V_2) \} \ne \emptyset $, and (ii) for every
non-collider structure ($ V_1 \rightarrow V_2 \rightarrow V_3 $, $ V_1
\leftarrow V_2 \rightarrow V_3 $), $ V_2 \not \in \bm{Z} $. $ X $ and $ Y $ are
d-separated given $ \bm{Z} $ if no d-connecting path exist between them.

@@

\begin{figure}
	\centering
	\begin{tabular}{llrl}
		$X$ & $Y$ & $ \bm{Z} $ & p-value \\
		\hline
		Incm & MrtS &  Age, Occp, Rltn & $ 0.29 $     \\
		Incm & Sex  &  Occp, Rltn      & $ 0.21 $     \\
		Incm & Wrkc &  Occp 	       & $ 1.5e^{-3} $ \\
		Edct & MrtS & Incm, Occp, Rltn & $ 0.02 $       \\
		\hline
	\end{tabular}
	\caption{Results of testing some of implied CIs of the DAG in
		 Figure~\ref{fig:intro}(b) using the same CI test that was used for
		 learning it.}
	\label{fig:ci_table}
\end{figure}

@@

whereas a \emph{weak ancestral oracle} is defined as:
$$
\mathcal{A}_G(X,Y)=\begin{cases}
 X \to Y & \textrm{if } X \in \textrm{An}_G(Y) \\
 X \gets Y & \textrm{if } Y \in \textrm{An}_G(X) \\
 \textrm{rand}(\{X \gets Y, X \to Y\}) & \textrm{otherwise} \\
\end{cases}
$$
where $\textrm{rand}(S)$ denotes a uniformly chosen element of $S$.
That is, our ancestral oracles only provide information on the direction of cause-effect
relationships, and do not distinguish between direct and indirect effects. 
The weak ancestral oracle is even unable to detect the absence of cause-effect
relationships between two variables. Using such oracles for DAG construction  
inherently risks that superfluous edges are created.

@@

In this section, we compare our \textsc{ExpertInLoop} algorithm with automated
causal discovery algorithms. Our goal with the empirical analysis is to
understand the behaviour of our method when the experts are not perfect. To
simulate an imperfect domain expert, we use a version of the strong ancestral oracle
that has an accuracy of $ \alpha $:
\begin{equation}
	\begin{split}
		x &= \textrm{rand}([0, 1]) \\
		\mathrm{Expert}(\alpha) &= \begin{cases} 
			\mathcal{O}(X, Y),  & \textrm{if  } x <= \alpha \\
			\textrm{rand}(X \rightarrow Y, Y \leftarrow X, \textrm{None}) & \textrm{otherwise} \\
				\end{cases} \\
			\mathcal{O}(X, Y) &= \begin{cases}
				X \rightarrow Y, & \textrm{if } X \in \textrm{An}_{G_{\textrm{true}}}(Y) \\
				X \leftarrow Y, & \textrm{if } Y \in \textrm{An}_{G_{\textrm{true}}}(X) \\
				\textrm{None}, & \textrm{otherwise }
				  \end{cases}
	\end{split}
\end{equation}
where $G_{\textrm{true}}$ is the true data generating DAG.

It is important to note that the effective accuracy of $ \mathrm{Expert} $ is
higher than $ \alpha $ as even when $ x > \alpha $, there is a $ 1 $ in
$ 3 $ chance that edge orientation is correct. Therefore, an $ \mathrm{Expert} $ with
accuracy $ \alpha $, has an effective accuracy, $ \alpha_{\mathrm{eff}} =
\alpha + (1 - \alpha) / 3 $.

@@

To compare the learned DAG to the original DAG, we use two metrics: Structural
Hamming Distance (SHD) and Structural Intervention Distance (SID)
\citep{Peters2015}. Unlike \textrm{ExpertInLoop}, the automated algorithms can
only recover the MEC, therefore, we considered all possible orientations of the
MEC to compute a range of SHD and SID values representing the best and worst
case scenarios. The results of this analysis are shown in
Figure~\ref{fig:shd_sid}. For SHD, the performance of \textsc{ExpertInLoop} is
comparable to the automated algorithms for $ \alpha = 0.3
(\alpha_{\textrm{eff}} = 0.53 $), implying that if an expert is able to get the
edge orientation correctly in more than $ 1 $ out of $ 2 $ cases, they can
outperform the automated algorithms. Similarly for SID, an expert can outperform 
the automated algorithms if they are able to get the orientation correct in $ 2 $ 
out of $ 3 $ cases. Another thing to note is that for denser DAGs, the 
\textsc{ExpertInLoop} is able to perform better for lower $ \alpha $ values.
